resevol: an R package for spatially explicit models of pesticide resistance given evolving pest genomes

### 2023-11-30 notebook/ms/plos_latex_ms.tex
Some edits to the manuscript prior to acceptance from PLoS Computational Biology

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={resevol: an R package for spatially explicit models of pesticide resistance given evolving pest genomes},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[R]{RESEVOL, AN R PACKAGE}
\usepackage{lineno}

@@

\modulolinenumbers[2]
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{acm}

\title{resevol: an R package for spatially explicit models of pesticide resistance given evolving pest genomes}
\author{A. Bradley Duthie\(^{1,*}\), Rosie Mangan\(^{1}\), C. Rose McKeon\(^{1}\),\\
Matthew C. Tinsley\(^{1}\), and Luc F. Bussière\(^{2,3}\)}
\date{{[}1{]} Biological and Environmental Sciences, University of Stirling, Stirling, UK {[}2{]}Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden {[}3{]} Gothenburgh Global Biodiversity Centre, Gothenburg, Sweden {[}*{]} \href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}}

\begin{document}
\maketitle

\setstretch{2}
\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{abstract}{%
\section*{Abstract}\label{abstract}}
\addcontentsline{toc}{section}{Abstract}

The evolution of pesticide resistance is a widespread problem with potentially severe consequences for global food security.
We introduce the resevol R package, which simulates individual-based models of pests with evolving genomes that produce complex, polygenic, and covarying traits affecting pest life history and pesticide resistance.
Simulations are modelled on a spatially-explicit and highly customisable landscape in which crop and pesticide application and rotation can vary, making the package a highly flexible tool for both general and tactical models of pest management and resistance evolution.
We present the key features of the resevol package and demonstrate its use for a simple example simulating pests with two covarying traits.
The resevol R package is open source under GNU Public License. All source code and documentation are available on GitHub.

\textbf{Key words:} \emph{pest management}, \emph{food security}, \emph{ecological modelling}, \emph{pesticide}, \emph{resistance evolution}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Insect resistance to pesticides is a wicked and widespread problem \citep{Gould2018}.
Predicting, identifying, and ultimately delaying the evolution of insecticide resistance is therefore critical for ensuring sustainable global food security \citep{Hawkins2019, Lykogianni2021}.
To manage pesticide resistance, various strategies have been proposed to disrupt or weaken selection.
These strategies often focus on varying the location and timing of pesticide application, and therefore varying the strength of selection for pesticide resistance, so that pest susceptibility to one or more pesticides is maintained \citep{Ives2002, Liang2013, Carriere2016, Liang2018, South2018, Sudo2018, Saikai2021}.
Pesticide resistance management has mostly been associated with the effect of single genes.
In such cases, resistance alleles have binary effects on phenotype, enabling resistant phenotypes to arise from genetic changes at single or a small number of loci \citep{Raymond2001, Ffrench-Constant2004, Ffrench-Constant2013}.
Nevertheless, resistance attributable to polygenic effects is also well-established \citep{Via1986, Firko1990, Busi2013, Haridas2018, Green2020, Mangan2023}.
And the relevance of polygenic resistance is likely to further increase given a rising interest in biological controls for sustainable crop protection \citep[e.g.,][]{Green2020, Elnahal2022, Mangan2023}.
Developing strategies to maximise the efficacy of these tools is critical, and such strategies should be well-informed by predictions made from detailed, quantitative genetic models.
Here we introduce the resevol R package as a tool for building individual-based models and simulating pest management \citep{Daly2022}.

@@

\begin{figure}
\centering
\includegraphics{ms_files/figure-latex/network-1.pdf}
\caption{\label{fig:network}Example network mapping loci (green circles) to traits (red diamonds) through an intermediate set of hidden layers (blue squares) in the \texttt{mine\_gmatrix} function. Individual genomes in the resevol R package consist of standard random normal values for loci, real values for black arrows linking nodes, and real values for traits. Values shown for loci and arrows are an example for illustration.}
\end{figure}

@@

An evolutionary algorithm is used to find appropriate values that produce covarying traits from loci (see Supporting Information 1 for details).
Evolutionary algorithms are heuristic tools that can simulate adaptive evolution to find solutions for a broad range of problems \citep{Hamblin2013, Duthie2018}.
In the resevol package, the \texttt{mine\_gmatrix} function runs an evolutionary algorithm and requires the argument \texttt{gmatrix}, which specifies the desired trait covariance matrix.
The function initialises a population of \texttt{npsize} separate, and potentially unique, networks (i.e., \texttt{npsize} copies of a network like the one shown in Fig \ref{fig:network}), and this population evolves until some maximum iteration (\texttt{max\_gen}) or minimum expected network stress (\texttt{term\_cri}) is met.
In a single iteration of the algorithm, values mutate and crossover occurs between networks (i.e., some of the \texttt{npsize} networks swap values with some probability).
Next, trait covariances produced for each network are estimated by initalising \texttt{indivs} individuals with loci sampled from a standard normal distribution.
Network stress is calculated as the logged mean squared deviation between estimated covariances and those in \texttt{gmatrix}.
Tournament selection \citep{Hamblin2013} is then used to determine the networks for the next iteration of the algorithm.
Networks with the lowest estimated stress have the highest fitness, so these networks are disproportionately represented in the next iteration.
Throughout the evolutionary algorithm, the lowest stress network is saved and returned upon network termination.
The robustness of this network's stress to sets of individuals with different loci values can be tested using the \texttt{stress\_test} function.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{trait\_covs  }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\AttributeTok{data =} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\SpecialCharTok{{-}}\FloatTok{0.4}\NormalTok{, }\SpecialCharTok{{-}}\FloatTok{0.4}\NormalTok{, }\DecValTok{1}\NormalTok{), }\AttributeTok{nrow =} \DecValTok{2}\NormalTok{, }\AttributeTok{ncol =} \DecValTok{2}\NormalTok{);}
\NormalTok{new\_network }\OtherTok{\textless{}{-}} \FunctionTok{mine\_gmatrix}\NormalTok{(}\AttributeTok{loci =} \DecValTok{3}\NormalTok{, }\AttributeTok{layers =} \DecValTok{2}\NormalTok{, }\AttributeTok{gmatrix =}\NormalTok{ trait\_covs, }
                            \AttributeTok{max\_gen =} \DecValTok{1000}\NormalTok{, }\AttributeTok{term\_cri =} \SpecialCharTok{{-}}\FloatTok{6.0}\NormalTok{, }\AttributeTok{prnt\_out =} \ConstantTok{FALSE}\NormalTok{);}
\end{Highlighting}
\end{Shaded}

@@

The mean deviation between elements of the above matrix and the identity matrix provided by \texttt{trait\_covs} is \ensuremath{2.1176024\times 10^{-4}}.
Lower values of \texttt{term\_cri} and higher values of \texttt{max\_gen} will result in a lower stress, but this will require more computation time, especially if the number of traits is high.
Similarly, higher values of \texttt{indivs} will result in more accurate estimations of true stress, but this also requires more computation time.
Additional arguments to \texttt{mine\_gmatrix} can also be used to improve the performance of the evolutionary algorithm (see Supporting Information 1).

@@

Landscapes are spatially explicit and initialised in one of two ways.
First, a landscape can be built from the arguments \texttt{xdim}, \texttt{ydim}, and \texttt{farms}.
These arguments specify the cell dimensions and number of farms on the landscape.
Contiguous rectangular farms of roughly equal size are generated on the landscape using a splitline algorithm.
Second, a custom landscape can be input using the \texttt{terrain} argument, which takes a matrix with elements that include integers 1 to \texttt{farms}.
Each value defines a unique farm, but values do not need to be contiguous.
These `farms' could even model non-farmland (e.g., water, roads), if pesticides and crops on them are invisible to pests (see Supporting Information 2), or they could more broadly be interpreted as heterogenous landscape properties across an arbitrary scale \citep[e.g.,][]{Newman2023}.
This \texttt{terrain} customisation therefore allows for a high degree of landscape detail, and offers the potential for modelling real-world landscapes from raster images \citep[e.g.,][]{Millington2021}.
Edge effects are set using the \texttt{land\_edge} argument.
Edge options include \texttt{"leaky"} (pests disappear from the landscape), \texttt{"reflect"} (pests bounce off of the edge), \texttt{"sticky"} (pests stick to the edge), or \texttt{"torus"} (pests that move off of one edge return on the opposite side of the landscape).

Each farm can hold one pesticide and one crop type in any time step.
The \texttt{crop\_init} and \texttt{pesticide\_init} arguments initialise one of \texttt{crop\_number} crops and one of \texttt{pesticide\_number} pesticides for each farm, respectively (maximum of 10 each).
Initialisation can be random for each farm with equal probability, or it can be set using a vector of length \texttt{farms} in which vector elements define the initialised crop or pesticide number.
After initialisation, crops and pesticides rotate once every \texttt{crop\_rotation\_time} and \texttt{pesticide\_rotation\_time} time steps, respectively.
The arguments \texttt{crop\_rotation\_type} and \texttt{pesticide\_rotation\_type} specify how crops and pesticides are rotated, respectively.
Each of these rotation type arguments can take either an integer value from 1-3, or a matrix.
Integer values specify (1) no rotation, (2) random transition from one type to another, or (3) cycling through each available crop or pesticide in numeric order.
Square matrices can be used to define the probability that a given crop or pesticide in row \(i\) transitions to that in column \(j\).
Hence, any possible Markov chain can be used to model transition between crop or pesticide types on farms, potentially integrating real crop use or pesticide use patterns \citep{Zhang2020}.
Upon rotation, crop and pesticide values are reset on each farm.

@@

In resevol v0.3, the \texttt{run\_farm\_sim} function includes 68 arguments, which specify a wide range of possible simulation conditions affecting landscape and pest characteristics.
These arguments are explained in the package documentation, and in Supporting Information 2, which demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
Here we focus on a simple simulation with asexually reproducing pests that have three loci and two traits (Fig \ref{fig:network}).
We use the the genome generated in \texttt{new\_network} from \protect\hyperlink{mine_gmatrix}{Section 2}, in which traits 1 and 2 have variances of 1.0008181 and 1.027134, respectively, with a covariance of -0.3925799.

>>>>

% Template for PLoS
% Version 3.6 Aug 2022
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}
%\usepackage{natbib}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage[nopatch=eqnum]{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{resevol: An R package for spatially explicit models of pesticide resistance given evolving pest genomes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
A. Bradley Duthie\textsuperscript{1*},
Rosie Mangan\textsuperscript{1},
C. Rose McKeon\textsuperscript{1},
Matthew C. Tinsley\textsuperscript{1},
Luc F. Bussi\`{e}re\textsuperscript{2,3}
\\
\bigskip
\textbf{1} Biological and Environmental Sciences, University of Stirling, Stirling, UK
\\
\textbf{2} Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
\\
\textbf{3} Gothenburgh Global Biodiversity Centre, Gothenburg, Sweden
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
%\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
%\ddag These authors also contributed equally to this work.

% Current address notes
%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
%\dag Deceased

% Group/Consortium Author Note
%\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* alexander.duthie@stir.ac.uk

\end{flushleft}

% Please keep the abstract below 300 words
\section*{Abstract}
The evolution of pesticide resistance is a widespread problem with potentially severe consequences for global food security.
We introduce the resevol R package, which simulates individual-based models of pests with evolving genomes that produce complex, polygenic, and covarying traits affecting pest life history and pesticide resistance.
Simulations are modelled on a spatially-explicit and highly customisable landscape in which crop and pesticide application and rotation can vary, making the package a highly flexible tool for both general and tactical models of pest management and resistance evolution.
We present the key features of the resevol package and demonstrate its use for a simple example simulating pests with two covarying traits.
The resevol R package is open source under GNU Public License. All source code and documentation are available on GitHub.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%\section*{Author summary}
%Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
Insect resistance to pesticides is a wicked and widespread problem {[}\protect\hyperlink{ref-Gould2018}{1}{]}.
Predicting, identifying, and ultimately delaying the evolution of insecticide resistance is therefore critical for ensuring sustainable global food security {[}\protect\hyperlink{ref-Hawkins2019}{2},\protect\hyperlink{ref-Lykogianni2021}{3}{]}.
To manage pesticide resistance, various strategies have been proposed to disrupt or weaken selection.
These strategies often focus on varying the location and timing of pesticide application, and therefore varying the strength of selection for pesticide resistance, so that pest susceptibility to one or more pesticides is maintained {[}\protect\hyperlink{ref-Ives2002}{4}--\protect\hyperlink{ref-Saikai2021}{10}{]}.
Pesticide resistance management has mostly been associated with the effect of single genes.
In such cases, resistance alleles have binary effects on phenotype, enabling resistant phenotypes to arise from genetic changes at single or a small number of loci {[}\protect\hyperlink{ref-Raymond2001}{11}--\protect\hyperlink{ref-Ffrench-Constant2013}{13}{]}.
Nevertheless, resistance attributable to polygenic effects is also well-established {[}\protect\hyperlink{ref-Via1986}{14}--\protect\hyperlink{ref-Mangan2023}{19}{]}.
And the relevance of polygenic resistance is likely to further increase given a rising interest in biological controls for sustainable crop protection {[}e.g., \protect\hyperlink{ref-Green2020}{18},\protect\hyperlink{ref-Mangan2023}{19},\protect\hyperlink{ref-Elnahal2022}{20}{]}.
Developing strategies to maximise the efficacy of these tools is critical, and such strategies should be well-informed by predictions made from detailed, quantitative genetic models.
Here we introduce the resevol R package as a tool for building individual-based models and simulating pest management {[}\protect\hyperlink{ref-Daly2022}{21}{]}.

@@

@@

\begin{figure}[!h]
\caption{{\bf Example resevol network.}
Example network mapping loci (green circles) to traits (red diamonds) through an intermediate set of hidden layers (blue squares) in the \texttt{mine\_gmatrix} function. Individual genomes in the resevol R package consist of standard random normal values for loci, real values for black arrows linking nodes, and real values for traits. Values shown for loci and arrows are an example for illustration.}
\label{fig:network}
\end{figure}

An evolutionary algorithm is used to find appropriate values that produce covarying traits from loci (see \nameref{S1_Text} for details).
Evolutionary algorithms are heuristic tools that can simulate adaptive evolution to find solutions for a broad range of problems {[}\protect\hyperlink{ref-Hamblin2013}{22},\protect\hyperlink{ref-Duthie2018}{23}{]}.
In the resevol package, the \texttt{mine\_gmatrix} function runs an evolutionary algorithm and requires the argument \texttt{gmatrix}, which specifies the desired trait covariance matrix.
The function initialises a population of \texttt{npsize} separate, and potentially unique, networks (i.e., \texttt{npsize} copies of a network like the one shown in Fig \ref{fig:network}), and this population evolves until some maximum iteration (\texttt{max\_gen}) or minimum expected network stress (\texttt{term\_cri}) is met.
In a single iteration of the algorithm, values mutate and crossover occurs between networks (i.e., some of the \texttt{npsize} networks swap values with some probability).
Next, trait covariances produced for each network are estimated by initalising \texttt{indivs} individuals with loci sampled from a standard normal distribution.
Network stress is calculated as the logged mean squared deviation between estimated covariances and those in \texttt{gmatrix}.
Tournament selection {[}\protect\hyperlink{ref-Hamblin2013}{22}{]} is then used to determine the networks for the next iteration of the algorithm.
Networks with the lowest estimated stress have the highest fitness, so these networks are disproportionately represented in the next iteration.
Throughout the evolutionary algorithm, the lowest stress network is saved and returned upon network termination.
The robustness of this network's stress to sets of individuals with different loci values can be tested using the \texttt{stress\_test} function.

@@

\begin{verbatim}
trait_covs  <- matrix(data = c(1, -0.4, -0.4, 1), nrow = 2, ncol = 2);
new_network <- mine_gmatrix(loci = 3, layers = 2, gmatrix = trait_covs, 
                            max_gen = 1000, term_cri = -6.0, 
                            prnt_out = FALSE);
\end{verbatim}

@@

The mean deviation between elements of the above matrix and the identity matrix provided by \texttt{trait\_covs} is \ensuremath{2.1176024\times 10^{-4}}.
Lower values of \texttt{term\_cri} and higher values of \texttt{max\_gen} will result in a lower stress, but this will require more computation time, especially if the number of traits is high.
Similarly, higher values of \texttt{indivs} will result in more accurate estimations of true stress, but this also requires more computation time.
Additional arguments to \texttt{mine\_gmatrix} can also be used to improve the performance of the evolutionary algorithm (see \nameref{S1_Text}).

@@

Landscapes are spatially explicit and initialised in one of two ways.
First, a landscape can be built from the arguments \texttt{xdim}, \texttt{ydim}, and \texttt{farms}.
These arguments specify the cell dimensions and number of farms on the landscape.
Contiguous rectangular farms of roughly equal size are generated on the landscape using a splitline algorithm.
Second, a custom landscape can be input using the \texttt{terrain} argument, which takes a matrix with elements that include integers 1 to \texttt{farms}.
Each value defines a unique farm, but values do not need to be contiguous.
These `farms' could even model non-farmland (e.g., water, roads), if pesticides and crops on them are invisible to pests (see \nameref{S2_Text}), or they could more broadly be interpreted as heterogenous landscape properties across an arbitrary scale {[}e.g., \protect\hyperlink{ref-Newman2023}{24}{]}.
This \texttt{terrain} customisation therefore allows for a high degree of landscape detail, and offers the potential for modelling real-world landscapes from raster images {[}e.g., \protect\hyperlink{ref-Millington2021}{25}{]}.
Edge effects are set using the \texttt{land\_edge} argument.
Edge options include \texttt{"leaky"} (pests disappear from the landscape), \texttt{"reflect"} (pests bounce off of the edge), \texttt{"sticky"} (pests stick to the edge), or \texttt{"torus"} (pests that move off of one edge return on the opposite side of the landscape).

Each farm can hold one pesticide and one crop type in any time step.
The \texttt{crop\_init} and \texttt{pesticide\_init} arguments initialise one of \texttt{crop\_number} crops and one of \texttt{pesticide\_number} pesticides for each farm, respectively (maximum of 10 each).
Initialisation can be random for each farm with equal probability, or it can be set using a vector of length \texttt{farms} in which vector elements define the initialised crop or pesticide number.
After initialisation, crops and pesticides rotate once every \texttt{crop\_rotation\_time} and \texttt{pesticide\_rotation\_time} time steps, respectively.
The arguments \texttt{crop\_rotation\_type} and \texttt{pesticide\_rotation\_type} specify how crops and pesticides are rotated, respectively.
Each of these rotation type arguments can take either an integer value from 1-3, or a matrix.
Integer values specify (1) no rotation, (2) random transition from one type to another, or (3) cycling through each available crop or pesticide in numeric order.
Square matrices can be used to define the probability that a given crop or pesticide in row \(i\) transitions to that in column \(j\).
Hence, any possible Markov chain can be used to model transition between crop or pesticide types on farms, potentially integrating real crop use or pesticide use patterns {[}\protect\hyperlink{ref-Zhang2020}{26}{]}.
Upon rotation, crop and pesticide values are reset on each farm.

@@

In resevol v0.3, the \texttt{run\_farm\_sim} function includes 68 arguments, which specify a wide range of possible simulation conditions affecting landscape and pest characteristics.
These arguments are explained in the package documentation, and in \nameref{S2_Text}, which demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
Here we focus on a simple simulation with asexually reproducing pests that have three loci and two traits (Fig \ref{fig:network}).
We use the the genome generated in \texttt{new\_network} from \protect\hyperlink{mine_gmatrix}{Section 1.1}, in which traits 1 and 2 have variances of 1.0008181 and 1.027134, respectively, with a covariance of -0.3925799.
